TBD

### 2020-09-06 Articles/2019-Entropy/Source/entropy-article.tex
Review with Christine

The overall goal is to describe a generic system as it evolves in time by defining a minimal set of concepts that such a setup must define, and understand the tacit assumptions underlying state spaces and evolution laws. We want to find notions that can be applied in different settings, being it thermodyamics and/or dynamical systems alike. Then we want to derive general results based on such definitions. Here we provide an overview of the main features, leaving the formal definitions in the (yet to be written) appendix.

@@

Our starting point is the \textbf{process domain} $\tdomain[P]$, which is the set that contains all possible descriptions of the system at all levels of detail at all possible times. Depending on the process, it will contain statements like \statement{the average volume between 1 and 2 seconds is between 3 and 4 liters} or \statement{the trajectory of the particle is $y=10 \, m - t^2 \cdot 9.80665 \, m/s^2 $}. It does not matter at this point what the statements themselves are, except they must be, at least in line of principle, defined from experimentally well-defined starting points. Mathematically, $\tdomain[P]$ is a $\sigma$-complete Boolean algebra\footnote{A set of statements that is closed under negation (NOT), countable disjunction (OR) and countable conjunction (AND).} and it is the closure of the set of experimentally verifiable statements $\edomain[P] \subseteq \tdomain[P]$.\footnote{The set $\edomain[P]$ will form a Heyting algebra which, to be experimentally reachable, must allow a countable basis. The closure $\tdomain[P]$ is in terms of negation and countable disjunction.}

As $\tdomain[P]$ contains different levels of description, some statement will be more or less specific. For example, we say that \statement{the horizontal position is between 2.5 and 3 meters} is \textbf{narrower} than \statement{the horizontal position is between 2 and 3.5 meters}. Note that whenever the first statement is true, the second one must be true as well. Given two statements $\stmt_1, \stmt_2 \in \tdomain[P]$, we formally capture this relationship by noting $\stmt_1 \narrower \stmt_2$. Narrowness can describe relationships at different scales, on different quantities (e.g. \statement{the horizontal position is between 2.5 and 3 meters and the vertical position is between 1 and 1.5 meters} $\narrower$ \statement{the horizontal position is between 2.5 and 3 meters}), constraints between variables at the same time (e.g. \statement{the temperature of the water in the glass is 3.98 C} $\narrower$ \statement{the density of the water in the glass is 1 $g/cm^3$}) or at different times (e.g. \statement{at time 0 s the position is 1 m and the velocity is 1 m/s} $\narrower$ \statement{at time 1 s the position is 2 m}). It is a general tool to capture relationships within the theory which intuitively can also be thought as implication.\footnote{Technically, implication in classical logic is a truth function as classical logic does not incorporate the idea of different ``possible cases''. What we have is more similar in spirit to semantic consequence in modal logic, though modal logic brings in a lot of undesirable elements we do not want.}

@@

Another logical relationship we want to capture is the idea of whether two statements can be true at the same time. For example, we want to say that \statement{the horizontal position is between 2.5 and 3.5 meters} is \textbf{compatible} with \statement{the horizontal position is between 2 and 3 meters}. In general, given two statements $\stmt_1, \stmt_2 \in \tdomain[P]$ we note $\stmt_1 \comp \stmt_2$ if it is possible for them to be both true. As with narrowness, compatibility can be in terms of different ranges of the same quantities, of different quantities and of constraints at equal or different times.

Finally, we want to capture whether the truth of one or more statements incluences the truth of others. For example, we want to say that \statement{the temperature of the water in the glass is 3.98 C} is \textbf{independent} from \statement{the volume of the water in the glass is 10 $cL$} but is not independent from \statement{the density of the water in the glass is 1 $g/cm^3$}. We note $\stmt_1 \indep \stmt_2$ when two statements are independent.

Every statement $\stmt \in \tdomain[P]$ can now be characterized by the set $A(\stmt) \subseteq E$ of all evolutions compatible with $\stmt$, that is all the evolutions for which $\stmt$ will be true. Mathematically,  $A(\edomain[P])$ is a \textbf{topology} over $E$ while $A(\tdomain[P])$ forms a $\sigma$-algebra over $E$, the \textbf{Borel algebra} of $A(\edomain[P])$. Narrowness and compatibility become set realtionships in the $\sigma$-algebra: $\stmt_1 \narrower \stmt_2$ if and only if $A(\stmt_1) \subseteq A(\stmt_2)$ (i.e.~all trajectories for which $\stmt_1$ is true are such that $\stmt_2$ is also true) and $\stmt_1 \comp \stmt_2$ if and only if $A(\stmt_1) \cap A(\stmt_2) \neq \emptyset$ (i.e.~two descriptions are compatible if there is an evolution that is compatbile with both). This connection gives a direct physical meaning to these foundational set-based mathematical structures.

\begin{table}[h!]
	\centering
\begin{tabular}[h]{|c|c|p{6cm}|}
	\hline 
	Symbol & Name & Meaning \\ 
	\hline 
	$\tdomain[P]$ & Process domain & The set of all possible descriptions at all times and at all levels of granularity \\ 
	\hline 
	$E \subset \tdomain[P]$ & Possible evolutions & The set of descriptions that give a full account of the process; they correspond to the narrowest statements within $\tdomain[P]$ \\ 
	\hline 
	$A(\stmt)$ & Compatible evolutions & The set of evolutions that are compatible with the statement $\stmt$ \\ 
	\hline 
	$\narrower$ & narrower & A statement is narrower than another $\stmt_1 \narrower \stmt_2$ if the second one is true whenever the first one is \\ 
	\hline 
	$\comp$ & compatible & A statement is compatible with another $\stmt_1 \comp \stmt_2$ if they can be true at the same time \\ 
	\hline 
	$\indep$ & independent & Two statements are independent  $\stmt_1 \comp \stmt_2$ if the truth of one does not influence the truth of the other \\ 
\hline 
\end{tabular} 
	\caption{Process descriptions and their logical relationships}
	\label{table:logic}
\end{table}

@@

As a process extends over time and potentially describes multiple systems, we want to organize the statements into partial domains, one for each moment in time relative to a particular system. Given a \textbf{time parameter} $t \in T \subseteq \mathbb{R}$, we can imagine to carve a \textbf{system domain} $\tdomain_t$ that would correspond to all the statements about the system one can make at the given time. In particular, we define a \textbf{one-step processes} one that only has an \textbf{initial domain} $\tdomain_{t_0}$ and a \textbf{final domain} $\tdomain_{t_1}$. Note that $\tdomain_t$ is a $\sigma$-complete Boolean sub-algebra of $\tdomain[P]$ since any logical operation on statements at fixed $t$ is also another statement at fixed $t$. The set $X_t \subset \tdomain_t$ of the narrowest statements will give the most precise description at that time, we call these the  possible \textbf{snapshots} of our system at that time. A \textbf{trajectory} is a sequence $x(t)$ of snapshots at each time. A particular evolution will identify a trajectory: that is, given $\stmt[e] \in E$ we can find $x(t)$ such that $\stmt[e] \narrower x(t_0) \AND x(t_1) \AND x(t_2) \AND ... $. The converse is not true: a single trajectory may correspond to multiple evolutions, since the evolutions potentially describe the same system at greater accuracy and/or other systems.

In fact, we should note that, technically, the statements are not defined at an infinitesimal moment $t$, rather within a finite interval $[t, t + \Delta t]$. This represents the time-scale at which the process will be described, meaning the description within $\tdomain_t$ should not be sensitive to what happens at faster scales. The quantities used to represent states at time $t$, then, should really be thought as averages within $\Delta t$. In the same way, $\tdomain_t$ will in general represent the system at the chosen level of description, not the narrowest level possible. Statements about the environment or at a finer level of description (i.e. the positions of all molecules for a gas) will not be included.

We should also note that the set of possible snapshots $X_t$ may not be the same at all times and may not correspond to the state of the system. First of all $X_t$ will only include those configurations that are compatible with at least by one possible evolution (i.e. $A(x) \neq \emptyset$ for all $x \in X_t$), which may change in time. For example, the set of possible configurations for a system under dissipative processes will shrink as they converge to equilibrium. In quantum mechanics, the possible configurations after a measurement are restricted to the eigenestates, and a different choice of measurement will lead to a different process with different possible evolutions. Secondly, the best possible description may be broader than the full state of the system (e.g. when external interference forces us to give only a statistical account within our $\Delta t$) or narrower (e.g. when there are known correlations to the other degrees of freedom of the environment or of the system itself). In other words, the state space of the system (i.e. the set of all possible complete descriptions of the system and only about that system, independently of the process at hand) is something we'll construct later.\footnote{In other words, starting directly from a state space comes with a set of implicit assumptions about the system which are at the foundations of thermodynamics specifically and physics more in general.}

@@

Conversely, we say a process is \textbf{reversible} over the system domains $\{\tdomain_t\}_{t \in T}$ if given the snapshot at one time we can always reconstruct the snapshot at all past times. Formally, let $t_1, t_2 \in T$ such that $t_1 \leq t_2$, for any $x_2 \in X_{t_2} \subset \tdomain_{t_2}$ we can find a $x_1 \in X_{t_1} \subset \tdomain_{t_1}$ such that $x_2 \narrower x_1$. In this case, all the evolutions that pass through $x_2$ must have passed through $x_1$. If $x(t)$ is the snapshot trajectory in time, we have $A(x(t)) \subseteq A(x(t - \Delta t))$: the set of compatible evolutions must stay the same or become smaller.

@@

If we limit ourselves to deterministic processes, with these few definitions, we have already found that the set of possible evolutions ``grows bigger'' under irreversible evolution while it remains the ``the same size'' under reversible evolution. This is the core idea of the second law.

@@

We want to stress that if $f$ is invertible, the system is \emph{not} necessarily reversible. In fact, we'd have $x_1 \equiv f^{-1}(x_2) \narrower x_2$, which is not the same as $x_2 \narrower g(x_2) \equiv x_1$: narrowness is in the opposite direction. In other words, \textbf{an invertible deterministic system is not necessarily reversible}. For example, consider a damped harmonic oscillator: the dynamics is in principle invertible (we can reconstruct the trajectory) but it is not reversible (as the evolutions bunch together, finite precision knowledge of the initial conditions gives us more information of the same finite precision knowledge of the final conditions). These types of distinctions are crucial we'd like to stress how it emerges naturally within the framework.

@@

Next step is to quantify the granularity of the description provided by each statement. Since statements at a finer level of description will put a greater constrain on the set of compatible evolutions, quantifying the granularity of the description of $\stmt$ means quantifying the size of $A(\stmt)$. The statement \statement{at time 0 s the particle is between 0 m and 1 m} is doubly more specific than \statement{at time 0 s the particle is between 0 m and 2 m} precisely because it correspond to half the evolutions of the second statement.

The naive approach would be to assign a measure over $A(\tdomain[P])$: since this is is a $\sigma$-algebra over $E$ and a measure is exactly one uses to give a size to each set, it would seem we are done. Unfortunately it is not that simple: a single measure can only compare objects of the same dimensionality. If we imagine all evolutions to be points on an $n$ dimensional manifold, if the measure gives finite values for $k$-dimensional regions it will necessarily give measure zero for all regions with lesser dimensionality and infinite measure for those with greater dimensionality. Coarse grained and subsystem descriptions will have different dimensionality as they describe different numbers of degrees of freedom, therefore we need a way to quantify objects at all levels.\footnote{Typically one uses geometrical structure for his purpose. In phase space, the symplectic form allows us to define areas of two-dimensional surfaces, and use those to define volumes or even-dimensional areas. Note that it does not quantify the size of the points, which are implicitly assumed to be of equal size. This is a special case and cannot be the starting point.}

Therefore our starting point will be the ability to compare two statements and decide which one gives a refined description of the process. Formally, we have partial order $\finer$ on $\tdomain[P]$ which, given two statements, it tells us whether the description of one is \textbf{finer}, more refined, than the other. Two statements are \textbf{equigranualar}, noted $\stmt_1 \eqgran \stmt_2$, if they provide the same level of description. For example, we can say that \statement{the trajectory of the particle is $y=t \cdot 1 \, m/s$} $\eqgran$ \statement{the trajectory of the particle is $y=t \cdot 2 \, m/s$} and that \statement{the position at time 0 sec is between 0 and 1 m} $\eqgran$ \statement{the position at time 0 sec is between 1 and 2 m}, even though the second pair is infinitely less discerning than the first. Note that if one statement is narrower than the other (i.e. $\stmt_1 \narrower \stmt_2$) then it is also finer than the other (i.e. $\stmt_1 \finer \stmt_2$) while the converse is not necessarily true.

We say two statements are \textbf{comparable} if one is finer than the other. The order is partial because not all statements are comparable to each other. For example, consider the pressure/volume state space for an ideal gas. Comparing \statement{the pressure is 1 kPa and the volume is between 1 and 2 liters} and \statement{the pressure is between 1 and 2 kPa and the volume is 1 liter} would mean saying where one unit liter is bigger or smaller than 1 kPa.\footnote{In phase space, for example, we cannot in general compare $q^i$ and $p_i$, yet we can always compare areas of phase space.}

The idea is that geometrical, measure theoretic, information theoretic and probabilistic structures all descend from this more fundamental structure, the partial order that describes this ``information granularity''. We note that all these structures can be used to determine a size and therefore a partial order on the set. It would be possible, given sufficient conditions, to recover those structure from the order itself. The mathematical detail of how this would exactly work (i.e. what condition are necessary and sufficient) is out of the scope of this work. Here we are interested in how the physical content attached to the partial order give rises to the other structure.

To recover measures, we pick a statement $\stmt[u] \in \tdomain[P]$ and construct a \textbf{measure} $\mu_{\stmt[u]}$ such that $\mu_{\stmt[u]}(\stmt[u]) = 1$ and $\stmt_1 \finer \stmt_2$ will mean $\mu_{\stmt[u]}(\stmt_1) \leq \mu_{\stmt[u]}(\stmt_2)$. That is, we can quantity how more or less precise a statement is compared to a fixed statement taken as a unit. If two statements are such that the measure of one with respect to the other is finite and non-zero, we say they are \textbf{finitely comparable}. We can show that if $\stmt[u]_1$ and $\stmt[u]_2$ are finitely comparable we have $ \mu_{\stmt[u]_1}(\stmt) = \mu_{\stmt[u]_1}(\stmt[u]_2) \mu_{\stmt[u]_2}(\stmt)$. Furthermore, since all evolutions give a complete description of the whole process, we will assume they are all equigranular. Therefore if we pick $\stmt[u] \in E$, then $\mu_{\stmt[u]}(\stmt)=\#(A(\stmt))$ would be the counting measure of evolutions compatible with $\stmt$.\footnote{We still needs to understand the necessary and sufficient conditions under which such measures would exist and be unique. For the present work, we assume it can be done. We know that it is a necessary condition that all evolution must be equigranular. It remains an open question is whether it is sufficient.}

We can also define the \textbf{evolution entropy} with respect to $\stmt[u]$ the quantity $S_{\stmt[u]}(\stmt) = \log \mu_{\stmt[u]}(\stmt)$. This can be thought as the number of bits, yes/no questions, that separate the level of description of $\stmt[u]$ from the one provided by $\stmt$. Connecting the definitions from above, under a deterministic process we have $S_{\stmt[u]}(x(t)) \leq S_{\stmt[u]}(x(t + \Delta t))$ since $x(t)\narrower x(t+\Delta t)$ means $x(t) \finer x(t+\Delta t)$ and $\mu_{\stmt[u]}(x(t)) \leq \mu_{\stmt[u]}(x(t+\Delta t))$. If the process is determistic and reversible, we have $S_{\stmt[u]}(x(t)) = S_{\stmt[u]}(x(t + \Delta t))$. In other words, \textbf{evolution entropy cannot decrease for a deterministic process and is stationary if the process is reversible as well}.

Quantifying evolutions allows us to define probability as well. If we consider all the evolutions compatible with \statement{the coin is fair}, we must have that half of them are compatible with \statement{the coin is head} while the other are compatible with \statement{the coin is tails}.  Therefore we define that the \textbf{probability} of $\stmt_2$ given $\stmt_1$ is
\begin{equation}
	P(\stmt_2 | \stmt_1) = \mu_{\stmt_1}(\stmt_1 \AND \stmt_2) = \frac{\mu_{\stmt[u]}(\stmt_1 \AND \stmt_2)}{\mu_{\stmt[u]}(\stmt_1)}
\end{equation} where $\stmt_1 \AND \stmt_2$ is the logical conjunction (AND) between the two statements and $\stmt[u]$ is finitely comparable to $\stmt_1$. In other words, the probability quantifies the ratio of evolutions compatible with $\stmt_1$ that are also compatible with $\stmt_2$. This can be shown to satisfy the standard axioms of probability.

Suppose that, given a process, we pick $t_1, t_2 \in T$. Given an initial description $\stmt_1 \in \tdomain_{t_1}$ and a final description $\stmt_2 \in \tdomain_{t_2}$, the \textbf{forward transition probability} $P(\stmt_2 | \stmt_1)$ tell us what is the fraction of evolutions starting in $\stmt_1$ (i.e. are compatible to $\stmt_1$)) that end in $\stmt_2$. In the same way, we can look at the \textbf{backwards transition probability}. The probability space is enough to recover the process, which means \textbf{forward and backward transition probabilities can always be used to fully characterize a process.}

>>>>

The overall goal is to describe a generic system as it evolves in time by defining a minimal set of concepts that such a setup must define, thus bringing to light the tacit assumptions underlying state spaces and evolution laws. We want to find notions that are equally valid and useful within different settings (e.g. thermodynamics, dynamical systems, classical mechanics, ...). We provide here an overview of the main features and a firm idea of how they recover the more standard results, leaving a complete formal development to future works.

@@

Our starting point is the \textbf{process domain} $\tdomain[P]$, which is the set that contains all possible descriptions of the system at all levels of detail at all possible times. Depending on the process, it will contain statements like \statement{the average volume of the system between 1 and 2 seconds is between 3 and 4 liters} or \statement{the trajectory of the particle is $y=10 \, m - t^2 \cdot 9.80665 \, m/s^2 $}. It does not matter at this point what the statements themselves are, except they must be, at least in line of principle, defined from experimentally well-defined starting points. The translation of these physical requirements into a formal structure has already been carried out in previous work, and we will briefly summarize them in this section. Mathematically, $\tdomain[P]$ is a $\sigma$-complete Boolean algebra\footnote{A set of statements that is closed under negation (NOT), countable disjunction (OR) and countable conjunction (AND).} and it is the closure of the set of experimentally verifiable statements $\edomain[P] \subseteq \tdomain[P]$.\footnote{The set $\edomain[P]$ will form a Heyting algebra which, to be experimentally reachable, must allow a countable basis. The closure $\tdomain[P]$ is in terms of negation and countable disjunction.}

As $\tdomain[P]$ contains different levels of description, some statements will be more or less specific. For example, we say that \statement{the horizontal position is between 2.5 and 3 meters} is \textbf{narrower} than \statement{the horizontal position is between 2 and 3.5 meters}. Note that whenever the first statement is true, the second one must be true as well. Given two statements $\stmt_1, \stmt_2 \in \tdomain[P]$, we formally capture this relationship by noting $\stmt_1 \narrower \stmt_2$ (read ``is narrower than''). Narrowness can describe relationships at different scales, on different quantities (e.g. \statement{the horizontal position is between 2.5 and 3 meters and the vertical position is between 1 and 1.5 meters} $\narrower$ \statement{the horizontal position is between 2.5 and 3 meters}), constraints between variables at the same time (e.g. \statement{the temperature of the water in the glass is 3.98 $^\circ C$} $\narrower$ \statement{the density of the water in the glass is 1 $g/cm^3$}) or at different times (e.g. \statement{at time 0 s the position is 1 m and the velocity is 1 m/s} $\narrower$ \statement{at time 1 s the position is 2 m}). It is a general tool to capture relationships within the theory which intuitively can also be thought of as implication.\footnote{Technically, implication in classical logic is a truth function as classical logic does not incorporate the idea of different ``possible cases''. What we have is more similar in spirit to semantic consequence in modal logic, though modal logic brings in a lot of undesirable elements we do not want.}

@@

Another logical relationship we want to capture is the idea of whether two statements can be true at the same time. For example, we want to say that \statement{the horizontal position is between 2.5 and 3.5 meters} is \textbf{compatible} with \statement{the horizontal position is between 2 and 3 meters}. In general, given two statements $\stmt_1, \stmt_2 \in \tdomain[P]$ we denote $\stmt_1 \comp \stmt_2$ (read ``is compatible with'') if it is possible for them to be both true. As with narrowness, compatibility can be in terms of different ranges of the same quantities, of different quantities and of constraints at equal or different times.

Finally, we want to capture whether the truth of one or more statements influences the truth of others. For example, we want to say that \statement{the temperature of the water in the glass is 3.98 $^\circ C$} is \textbf{independent} from \statement{the volume of the water in the glass is 10 $cL$} but is not independent from \statement{the density of the water in the glass is 1 $g/cm^3$}. We note $\stmt_1 \indep \stmt_2$ (read ``is independent of'') when two statements are independent.

Every statement $\stmt \in \tdomain[P]$ can now be characterized by the set $A(\stmt) \subseteq E$ of all evolutions compatible with $\stmt$, that is all the evolutions for which $\stmt$ will be true. Mathematically,  $A(\edomain[P])$ is a \textbf{topology} over $E$ while $A(\tdomain[P])$ forms a $\sigma$-algebra over $E$, the \textbf{Borel algebra} of $A(\edomain[P])$. Narrowness and compatibility become set relationships in the $\sigma$-algebra: $\stmt_1 \narrower \stmt_2$ if and only if $A(\stmt_1) \subseteq A(\stmt_2)$ (i.e.~all evolutions for which $\stmt_1$ is true are such that $\stmt_2$ is also true) and $\stmt_1 \comp \stmt_2$ if and only if $A(\stmt_1) \cap A(\stmt_2) \neq \emptyset$ (i.e.~two descriptions are compatible if there is an evolution that is compatible with both). This connection gives a direct physical meaning to these foundational set-based mathematical structures.

\begin{table}[h!]
	\centering
\begin{tabular}[h]{|c|c|p{6cm}|}
	\hline 
	Symbol & Name & Meaning \\ 
	\hline 
	$\tdomain[P]$ & Process domain & The set of all possible descriptions at all times and at all levels of granularity \\ 
	\hline 
	$E \subset \tdomain[P]$ & Possible evolutions & The set of descriptions that give a full account of the process; they correspond to the narrowest statements within $\tdomain[P]$ \\ 
	\hline 
	$A(\stmt)$ & Compatible evolutions & The set of evolutions that are compatible with the statement $\stmt$ \\ 
	\hline 
	$\narrower$ & narrower & A statement is narrower than another $\stmt_1 \narrower \stmt_2$ if the second one is true whenever the first one is \\ 
	\hline 
	$\comp$ & compatible & A statement is compatible with another $\stmt_1 \comp \stmt_2$ if they can be true at the same time \\ 
	\hline 
	$\indep$ & independent & Two statements are independent  $\stmt_1 \indep \stmt_2$ if the truth of one does not influence the truth of the other \\ 
\hline 
\end{tabular} 
	\caption{Process descriptions and their logical relationships}
	\label{table:logic}
\end{table}

@@

As a process extends over time and potentially describes multiple systems, we want to organize the statements into partial domains, one for each moment in time relative to a particular system. Given a \textbf{time parameter} $t \in T \subseteq \mathbb{R}$, we can imagine carving out a \textbf{system domain} $\tdomain_t$ that would correspond to all the statements about the system one can make at the given time. In particular, we define \textbf{one-step process} as one that only has an \textbf{initial domain} $\tdomain_{t_0}$ and a \textbf{final domain} $\tdomain_{t_1}$. Note that $\tdomain_t$ is a $\sigma$-complete Boolean sub-algebra of $\tdomain[P]$ since any logical operation on statements at fixed $t$ is also another statement at fixed $t$. The set $X_t \subset \tdomain_t$ of the narrowest statements will give the most precise description at that time; we call these the  possible \textbf{snapshots} of our system at that time. A \textbf{trajectory} is a sequence $x(t)$ of snapshots at each time. A particular evolution will identify a trajectory: that is, given $\stmt[e] \in E$ we can find $x(t)$ such that $\stmt[e] \narrower x(t_0) \AND x(t_1) \AND x(t_2) \AND ... $. The converse is not true: a single trajectory may correspond to multiple evolutions, since the evolutions potentially describe the same system at greater accuracy and/or other systems.

In fact, we should note that, technically, the statements are not defined at an infinitesimal moment $t$, but rather within a finite interval $[t, t + \Delta t]$. This represents the time-scale at which the process will be described, meaning the description within $\tdomain_t$ should not be sensitive to what happens at faster scales. The quantities used to represent states at time $t$, then, should really be thought of as averages within $\Delta t$. In the same way, $\tdomain_t$ will in general represent the system at the chosen level of description, not the narrowest level possible. Statements about the environment or at a finer level of description (i.e. the positions of all molecules for a gas) will not be included.

We should also note that the set of possible snapshots $X_t$ may not be the same at all times and may not correspond to the state of the system. First of all $X_t$ will only include those configurations that are compatible with at least one possible evolution (i.e. $A(x) \neq \emptyset$ for all $x \in X_t$), which may change in time. For example, the set of possible configurations for a system under dissipative processes will shrink as they converge to equilibrium. In quantum mechanics, the possible configurations after a measurement are restricted to the eigenstates, and a different choice of measurement will lead to a different process with different possible evolutions. Secondly, the best possible description may be broader than the full state of the system (e.g. when external interference forces us to give only a statistical account within our $\Delta t$) or narrower (e.g. when there are known correlations to the other degrees of freedom of the environment or of the system itself). In other words, the state space of the system (i.e. the set of all possible complete descriptions of the system and only about that system, independently of the process at hand) is something we'll construct later.\footnote{In other words, starting directly from a state space comes with a set of implicit assumptions about the system which are at the foundations of thermodynamics specifically and physics more in general.}

@@

Conversely, we say a process is \textbf{reversible} over the system domains $\{\tdomain_t\}_{t \in T}$ if given the snapshot at one time we can always reconstruct the snapshot at all past times. Formally, let $t_1, t_2 \in T$ such that $t_1 \leq t_2$, for any $x_2 \in X_{t_2} \subset \tdomain_{t_2}$ we can find an $x_1 \in X_{t_1} \subset \tdomain_{t_1}$ such that $x_2 \narrower x_1$. In this case, all the evolutions that pass through $x_2$ must have passed through $x_1$. If $x(t)$ is the snapshot trajectory in time, we have $A(x(t)) \subseteq A(x(t - \Delta t))$: the set of compatible evolutions must stay the same or become smaller.

@@

If we limit ourselves to deterministic processes, with these few definitions, we have already found that the set of possible evolutions ``grows bigger'' under irreversible evolution while it remains the ``the same size'' under reversible evolution. This is the core idea that will recover the second law of thermodynamics.

@@

We want to stress that if $f$ is invertible, the system is \emph{not} necessarily reversible. In fact, we would have $x_1 \equiv f^{-1}(x_2) \narrower x_2$, which is not the same as $x_2 \narrower g(x_2) \equiv x_1$: narrowness is in the opposite direction. In other words, \textbf{an invertible deterministic system is not necessarily reversible}. For example, consider a damped harmonic oscillator: the dynamics is in principle invertible (we can reconstruct the trajectory) but it is not reversible (as the evolutions bunch together, finite precision knowledge of the initial conditions gives us more information than the same finite precision knowledge of the final conditions). These types of distinctions are crucial and they emerge naturally within the framework.

@@

The next step is to quantify the granularity of the description provided by each statement. Since statements at a finer level of description will put a greater constraint on the set of compatible evolutions, quantifying the granularity of the description of $\stmt$ means quantifying the size of $A(\stmt)$. The statement \statement{at time 0 s the particle is between 0 m and 1 m} is doubly more specific than \statement{at time 0 s the particle is between 0 m and 2 m} precisely because it corresponds to half the evolutions of the second statement.

The naive approach would be to assign a measure over $A(\tdomain[P])$: since this is a $\sigma$-algebra over $E$ and a measure is exactly what one uses to give a size to each set, it would seem we are done. Unfortunately it is not that simple: a single measure can only compare objects of the same dimensionality. If we imagine all evolutions to be points on an $n$-dimensional manifold, if the measure gives finite values for $k$-dimensional regions it will necessarily give measure zero for all regions with lesser dimensionality and infinite measure for those with greater dimensionality. Coarse grained and subsystem descriptions will have different dimensionality as they describe different numbers of degrees of freedom, therefore we need a way to quantify objects at all levels.\footnote{Typically one uses geometrical structures for his purpose. In phase space, the symplectic form allows us to define areas of two-dimensional surfaces, and use those to define volumes or even-dimensional areas. Note that it does not quantify the size of the points, which are implicitly assumed to be of equal size. This is a special case and cannot be the starting point.}

Therefore our starting point will be the ability to compare two statements and decide which one gives a finer description of the process. Formally, we have partial order $\finer$ on $\tdomain[P]$ which, given two statements, tells us whether the description of one is \textbf{finer}, more refined, than the other. Two statements are \textbf{equigranualar}, noted $\stmt_1 \eqgran \stmt_2$, if they provide the same level of description. For example, we can say that \statement{the trajectory of the particle is $y=t \cdot 1 \, m/s$} $\eqgran$ \statement{the trajectory of the particle is $y=t \cdot 2 \, m/s$} and that \statement{the position at time 0 sec is between 0 and 1 m} $\eqgran$ \statement{the position at time 0 sec is between 1 and 2 m}, even though the second pair is infinitely less discerning than the first. Note that if one statement is narrower than the other (i.e. $\stmt_1 \narrower \stmt_2$) then it is also finer than the other (i.e. $\stmt_1 \finer \stmt_2$) while the converse is not necessarily true.

We say two statements are \textbf{comparable} if one is finer than the other. The order $\finer$ is partial because not all statements are comparable to each other. For example, consider the pressure/volume state space for an ideal gas. Comparing \statement{the pressure is 1 kPa and the volume is between 1 and 2 liters} and \statement{the pressure is between 1 and 2 kPa and the volume is 1 liter} would mean saying whether one liter is bigger or smaller than one kPa.\footnote{In phase space, for example, we cannot in general compare $q^i$ and $p_i$, yet we can always compare areas of phase space} The idea of physical dimension would therefore be something that is formally captured mathematically.

The idea is that geometrical, measure theoretic, information theoretic and probabilistic structures all descend from this more fundamental structure, the partial order that describes this ``information granularity''. We note that all these structures can be used to determine a size and therefore a partial order on the set. It would be possible, given sufficient conditions, to recover those structures from the order itself. The mathematical detail of how exactly this would work (i.e. what conditions are necessary and sufficient) is out of the scope of this work. Here we are interested in how the physical content attached to the partial order gives rise to the other structures.

To recover measures, we pick a statement $\stmt[u] \in \tdomain[P]$, called unit, and construct a \textbf{measure} $\mu_{\stmt[u]}$ such that $\mu_{\stmt[u]}(\stmt[u]) = 1$ and $\stmt_1 \finer \stmt_2$ will mean $\mu_{\stmt[u]}(\stmt_1) \leq \mu_{\stmt[u]}(\stmt_2)$. That is, we can quantify how much more or less precise a statement is compared to a fixed statement taken as a unit. If two statements are such that the measure of one with respect to the other is finite and non-zero, we say they are \textbf{finitely comparable}. We can show that if $\stmt[u]_1$ and $\stmt[u]_2$ are finitely comparable we have $ \mu_{\stmt[u]_1}(\stmt) = \mu_{\stmt[u]_1}(\stmt[u]_2) \mu_{\stmt[u]_2}(\stmt)$. Furthermore, since all evolutions give a complete description of the whole process, we will assume they are all equigranular. Therefore if we pick $\stmt[u] \in E$, then $\mu_{\stmt[u]}(\stmt)=\#(A(\stmt))$ would be the counting measure of evolutions compatible with $\stmt$.\footnote{We still need to understand the necessary and sufficient conditions under which such measures would exist and be unique. For the present work, we assume it can be done.}

We can also define the \textbf{evolution entropy} with respect to $\stmt[u]$ as the quantity $S_{\stmt[u]}(\stmt) = \log \mu_{\stmt[u]}(\stmt)$. This can be thought of as the number of bits, yes/no questions, that separate the level of description of $\stmt[u]$ from the one provided by $\stmt$. Connecting the definitions from above, under a deterministic process we have $S_{\stmt[u]}(x(t)) \leq S_{\stmt[u]}(x(t + \Delta t))$ since $x(t)\narrower x(t+\Delta t)$ means $x(t) \finer x(t+\Delta t)$ and $\mu_{\stmt[u]}(x(t)) \leq \mu_{\stmt[u]}(x(t+\Delta t))$. If the process is determistic and reversible, we have $S_{\stmt[u]}(x(t)) = S_{\stmt[u]}(x(t + \Delta t))$. In other words, \textbf{evolution entropy cannot decrease for a deterministic process and is stationary if the process is reversible as well}.

Quantifying evolutions allows us to define probability. The idea is a process can be realized differently by each possible evolution. The more evolutions a statement is compatible with, the more it is likely to happen. For example, consider a process that implements a coin flip. If the coin (and the process flipping the coin) is fair, we must have that half the evolutions are compatible with \statement{the coin is head} while the other half are compatible with \statement{the coin is tails}.  Therefore we define that the \textbf{probability} of $\stmt_2$ given $\stmt_1$ is
\begin{equation}
	P(\stmt_2 | \stmt_1) = \mu_{\stmt_1}(\stmt_1 \AND \stmt_2) = \frac{\mu_{\stmt[u]}(\stmt_1 \AND \stmt_2)}{\mu_{\stmt[u]}(\stmt_1)}
\end{equation} where $\stmt_1 \AND \stmt_2$ is the logical conjunction (AND) between the two statements and $\stmt[u]$ is finitely comparable to $\stmt_1$. In other words, the probability quantifies the ratio of evolutions compatible with $\stmt_1$ that are also compatible with $\stmt_2$. This can be shown to satisfy the standard axioms of probability.

Suppose that, given a process, we pick $t_1, t_2 \in T$. Given an initial description $\stmt_1 \in \tdomain_{t_1}$ and a final description $\stmt_2 \in \tdomain_{t_2}$, the \textbf{forward transition probability} $P(\stmt_2 | \stmt_1)$ tells us the fraction of evolutions starting in $\stmt_1$ (i.e. that are compatible with $\stmt_1$) that end in $\stmt_2$. In the same way, we can look at the \textbf{backward transition probability}. The transition probabilities are fully determined, through the measure, by the fineness relationship. Conversely, the fineness relationship can be, through the measure, recovered by the probabilities. Therefore \textbf{forward and backward transition probabilities are enough to fully characterize the behavior of a system within a process.}

We say that a process is \textbf{isentropic} if for any possible trajectory $x(t)$ we have $S_{\stmt[u]}(x(t)) = S_{\stmt[u]}(x(t + \Delta t)) \neq 0$. We saw that all deterministic and reversible processes are isentropic, but the converse is not true. We can give, though, a characterization in terms of transition probabilities. Assume $\stmt[u]$, $x(t)$ and $x(t+\Delta t)$ are all finitely comparable. We can write:
\begin{align*}
	\mu_{\stmt[u]}(x(t) &\AND x(t+\Delta t)) \\
	&=  \mu_{\stmt[u]}(x(t+\Delta t)) P(x(t)|x(t+\Delta t)) \\
	&=  \mu_{\stmt[u]}(x(t)) P(x(t+\Delta t)|x(t)) \\
\end{align*}
Therefore
\begin{equation}
\frac{\mu_{\stmt[u]}(x(t+\Delta t))}{\mu_{\stmt[u]}(x(t))} = \frac{P(x(t+\Delta t)|x(t))}{P(x(t)|x(t+\Delta t))}.
\end{equation}
If the forward and backward probabilities are the same, the initial and final snapshots have the same measure. Conversely, if the initial and final snapshot have the same measure, the forward and backward probabilities must be the same. Equal measure corresponds to equal process entropy. Therefore \textbf{isentropic processes are those and only those for which backward and forward probabilities are the same}.
